Thesis: draft

### 2021-05-30 draft.tex
latest update

This chapter is subdivided into four main sections. We first discuss the impact of network structure on the evolution of the prevalence of the opinions and the formation of echo chambers. Then, we introduce stubbornness in the system. The impact of how the opinions are initially distributed in a network with community structure (SBM) will be discussed afterwards. Finally, we will study the opinion dynamics model in real-world networks. We will discuss the results and compare them to results from network models with a similar average degree, clustering or modularity.\\
\newline
Four different network models are studied: (i) Erd\H{o}s-R\'{e}nyi model (ER); (ii) Watts-Strogatz model (WS); (iii) stochastic block model (SBM) and (iv) stochastic Watts-Strogatz block model (SBM-WS). Each model has $N = 10^3$ nodes and average degree $\left<k\right> \sim 10$. The ER model has an edge probability $p = 0.01$. For the WS model, we use two rewiring probabilities: $\beta = 0.01$ ($\left<cc\right> = 0.648 \pm 0.003$) and $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). The SBM has 10 communities of size 100 nodes each. We study one configuration with high modularity ($Q = 0.908 \pm 0.005$, with an edge probability inside the communities $p_{\text{cl}} = 0.1$ and an edge probability between communities $p_{\text{add}} = 0.001$) and one configuration with low modularity ($Q = 0.220 \pm 0.007$, with $p_{\text{cl}} = 0.03$ and $p_{\text{add}} = 0.008$). The SBM-WS has 10 communities of size 100 nodes each. Each community has WS parameters: mean degree $\left<k\right> =10$ and rewire probability $\beta = 0.01$. Edges are added between communities with a probability $p_{\text{add}} = 0.001$ ($\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$).\\
\newline
The opinion dynamics model discussed in Section \ref{modelThesis} is used. We set the activation probability equal to $p_{\text{act}} = 0.1$ and the actual time-line of each actor has a maximum length equal to $S=20$. If another model is used or if parameters (either those of the network models or the opinion dynamics model) are changed, this will be explicitly stated in the relevant sections. Opinions and stubborns (if present) are randomly distributed over the network, unless specified otherwise.

@@

Real-world social networks often have high clustering and modular structure. We want to study the impact of these topologies on the dynamics of opinions. In this section, we will discuss results related to different network topologies with focus on the effect of clustering and community structure. Two different opinion distributions are investigated, balanced and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions and the formation of echo chambers. Two filtering algorithms will be compared, the REC and the PR filtering algorithms, in order to investigate the interplay between filtering algorithms and network structure.

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_REC_8x7.png}
      \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_PR_8x7.png}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$, for balanced starting conditions.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$, for balanced starting conditions. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Four network models are used: (i) ER; (ii) SBM; (iii) SBM-WS and (iv) WS. Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. For the WS model, we use two different values for the rewire probability: $\beta = 0.01$ ($\left<cc\right> = 0.648 \pm 0.003$; WS2) and $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$; WS1). The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line represents a constant amount of echo chambers over time. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}
\label{echo_50_50}
\end{figure}
\noindent
In Fig. \ref{echo_50_50} two different filtering algorithms are compared, the REC and the PR filtering algorithms. The results for both filtering algorithms are qualitatively the same. The difference is that the formation of echo chambers, using the PR filtering algorithm, is about ten times as large compared to the formation of echo chambers using the REC filtering algorithm. \\
\newline
The ER model shows no, or a limited, sign of echo chambers, for both the REC and PR filtering algorithms. Networks with high clustering (WS and SBM-WS), on the other hand, show a sign of polarization and formation of echo chambers (visible in the two spikes, one around zero and another around one). Going to slightly higher values of the clustering coefficient ($\left<cc\right> = 0.557 \pm 0.007$ to $\left<cc\right> = 0.648 \pm 0.003$) in the WS model results in significantly larger echo chambers. This suggests that the formation of echo chambers is sensitive to changes in the clustering coefficient. The SBM with high modularity, community structure, shows an appearance of polarization (mainly when using the PR filtering algorithm), however, the formation of echo chambers is significantly smaller compared to networks with high clustering. The stochastic block model with low modularity has no, or limited, formation of echo chambers, similar as for the ER model. This is expected, since a SBM with low modularity, and hence, almost non-existing community structure, resembles an ER network. The SBM-WS has both community structure and high average clustering. The formation of echo chambers when using this model is approximately of the same size as the formation of echo chambers in a WS model with similar average clustering. This suggests that it is the high clustering that drives the dynamics to form echo chambers, while the community structure plays a less important role. \\
\newline
For balanced starting conditions, none of the network models, nor the filtering algorithms could break the status quo (Fig. \ref{ev_op_50_50} and related results in \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can potentially shift the initial randomly distributed opinions to a clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

\begin{figure}[H]
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_001_50-50_PR_REC_20-80_8x7.png}
      \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10-006_PR_REC_20-80_8x7.png}
    \label{ws_20-80_op}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}    
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_01-0001_003-0008_PR_REC_20-80_8x7.png}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10-001-0001_PR_REC_20-80_8x7.png}
    \label{sbm-ws_20-80_op}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$ configuration). For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_20_80}
\end{figure}
\noindent
This result differs from results in \cite{Perra2019}, where the minority opinion was able to survive over time for all the network models, even for the PR filtering algorithm, and where a minority opinion prevalence at $t=500$ between $P_\text{A}(500) = 0.06$ and $P_\text{A}(500) = 0.09$ was reported \cite{Perra2019}. The difference in the results is most likely due to an interplay between the network size $N$, the average degree $\left<k\right>$ and the maximum length of the actual time-line $S$. The results that where obtained in \cite{Perra2019} are for networks with size $N = 10^5$ nodes and average degree $\left<k\right> \sim 6$. In this thesis results are obtained for $N=10^3$ and average degree $\left<k\right> \sim 10$. In both cases the maximum length of the actual time-line is $S=20$. Fig. \ref{ev_op_WS_k=6} shows the results for the evolution of the prevalence of opinion A for an initial $20/80$ opinion distribution and using the PR filtering algorithm; Fig. \ref{ws1000} displays results for a WS model with $N=10^3$ and $\left<k\right> = 6$; Fig. \ref{ws10000} shows results for a WS model with $N=10^4$ and $\left<k\right> = 6$. We see that the minority opinion A reaches higher final values of $P_0(500)$ compared to the case depicted in Fig. \ref{ws_20-80_op}, with a value of $P_\text{A}(500) = 0.027 \pm 0.023$ in Fig. \ref{ws1000} and a value of $P_\text{A}(500) = 0.030 \pm 0.018$ in Fig. \ref{ws10000}.

@@

  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_\text{B}^{\text{nn}}\right>$, for unbalanced starting conditions.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$ for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. For each model the REC and the PR filtering algorithms are compared. The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line represents a constant amount of echo chambers over time. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}

@@

Lets first discuss the results for the ER model and the SBM (Fig. \ref{er_echo_20-80} and Fig. \ref{sbm}). When using the REC filtering algorithm, the formation of echo chambers around the majority opinion B is hampered. This is probably a consequence of the absence of clustering and the presence of short average path lengths in these models. There is even a little spike around $\left<P_\text{B}^{\text{nn}}\right> = 0.1$, which indicates a small increase in the fraction of actors with the majority of their neighbors carrying the minority opinion A. From Fig. \ref{ev_op_20_80} it was concluded that the REC filtering algorithm does not break the initial prevalence of the opinions. The opinion distribution is however slightly shifted from the case where the minority opinion A is randomly distributed over the network to the case where the minority opinion A starts to group together in opinion clusters. \\

@@

The PR filtering algorithm shows a different tendency. There is an increase in the fraction of actors with (almost) all their neighbors carrying the majority opinion B. This is a consequence from the fact that the PR filtering algorithm reduces the visibility of the minority opinion A in the neighborhood of actors with the majority opinion B. This reduced visibility of the minority opinion A in these neighborhoods drives them eventually to adopt opinion B, with an increase in the number of echo chambers around the majority opinion B as a consequence. Remember from Fig. \ref{ev_op_20_80}, that the PR filtering algorithm reduces the initial prevalence of the minority opinion. Hence, the increase in the formation of echo chambers around the majority opinion B is not only a consequence of a reorganization of the system from a random opinion distribution to a clustered opinion distribution, but also from a shift in the initial $20/80$ opinion distribution to a final (close to) $0/100$ opinion distribution. This can also be seen in the fact that, across the $x$-axis, the average number of neighborhoods with some fraction of the actors carrying the minority opinion A becomes equal to zero. \\

@@

Networks with high clustering (WS and SBM-WS) behave completely different than those with absence of clustering. For both the REC and PR filtering algorithms the high clustering enhances the formation of echo chambers around the minority opinion A (visible in the spike around $\left<P_\text{B}^{\text{nn}}\right> = 0$ in Fig. \ref{ws} and Fig. \ref{sbm-ws}). The absolute value of neighborhoods with all the actors carrying the minority opinion A at $t=0$ is equal to zero in the case of a WS network in Fig. \ref{ws} and in the case of a SBM-WS network in Fig. \ref{sbm-ws}. This leads to a normalized average value of the echo chambers, which is obtained by dividing the value at $t=500$ by the corresponding value at $t=0$, equal to infinity. In Fig. \ref{ws} and Fig. \ref{sbm-ws}, the spikes around zero are not actually reaching infinity, but are cut off around values equal to 100/180. These large spikes, indicating a large increase of echo chambers around the minority opinion, do not necessary mean that we have a large absolute number of echo chambers around the minority opinion A at $t=500$. They merely indicate that, compared to the beginning at $t=0$, the increase is large (and if there are zero echo chambers around the minority opinion A at $t=0$, the increase would be infinite even if there is only one echo chamber around the minority opinion A at $t=500$). If we would look at the absolute values of the number of echo chambers at $t=500$, we would still find that the vast majority of the like-minded neighborhoods carry the majority opinion B and that only a minor fraction carries the minority opinion A.\\

@@

\begin{table}[H]
\centering
\setlength{\tabcolsep}{10pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1
\begin{tabular}{c | c}
$\bm{Q}$ & \textbf{Network Model} \\
\hline
$0$ & ER model; $p=0.01$ \\
$0.220 \pm 0.007$ & SBM; $p_{\text{cl}} = 0.03$, $p_{\text{add}} = 0.008$ \\
$0.476 \pm 0.009$ & SBM; $p_{\text{cl}} = 0.05$, $p_{\text{add}} = 0.005$ \\
$0.624 \pm 0.008$ & SBM; $p_{\text{cl}} = 0.07$, $p_{\text{add}} = 0.004$ \\
$0.908 \pm 0.005$ & SBM; $p_{\text{cl}} = 0.1$, $p_{\text{add}} = 0.001$ \\
\end{tabular}
\caption{Network models used in Fig. \ref{echo_vs_mod}.}
\label{tab1}
\end{table}
\noindent
For most values of the modularity, the REC filtering algorithm produces almost no echo chambers. If the modularity becomes high enough we see a small increase in the formation of echo chambers, however even for $Q = 0.908 \pm 0.005$, the size of the echo chambers remains smaller than two. This means that after 500 time steps there are less than twice as much echo chambers than in the beginning at $t=0$. For the PR filtering algorithm, the increase in the size of the echo chambers with increasing modularity is more significant. From a modularity equal to $Q = 0.476 \pm 0.009$, we see a sharp increase in the formation of echo chambers. For $Q < 0.476$, the formation of echo chambers remains small, even when using the PR filtering algorithm. These results suggest that community structure enhances the formation of echo chambers, mostly when using the PR filtering algorithm. High modularity is however required to see the effect of community structure on the formation of opinion clusters.
\newpage
\begin{figure}[H]
	\includegraphics[width=10cm, height=8cm]{echo_chamber_vs_modularity_edge_prob_same_deg_PR_REC_10x8}
	\captionsetup{format=plain}
	\caption[Relation between the formation of echo chambers and community structure (modularity $Q$).]{Relation between the formation of echo chambers and the community structure, measured by the modularity $Q$, for both the PR and REC filtering algorithms. Data points with $Q = 0$ are the result for an ER network; the other eight data points are results for SBM's, each with different modularity, which is changed by changing the edge probabilities. The dotted black line represents a constant amount of echo chambers over time. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B  (normalized means that the value for the fraction of actors with all neighbors having opinion A or all neighbors having opinion B at $t=500$ is divided by the corresponding value at $t=0$). Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$. The data points are shown together with the standard error $SE = \sigma /\sqrt{n}$ ($n$ is the number of independent simulations), because we are most interested in the correctness of the average.}
\label{echo_vs_mod}
\end{figure}
\noindent
\section{Stubborn actors}

Stubbornness is an important feature of real-life opinion dynamics. Understanding the impact of individual resistance and the interplay with filtering algorithms and network topologies is crucial. In this section, we will study the effect of stubborn actors on the opinion dynamics. We focus on two different stubbornness models, discussed in Section \ref{stubbBin}. The results of the first model, including a resistance parameter, are given in Section \ref{stubbpar}. The results of the second model, the majority threshold model, are presented in Section \ref{majThres}.\\
\newline
We will compare the REF, REC and PR filtering algorithms to analyze the interplay between stubbornness and filtering algorithm. The ER, WS, SBM and SBM-WS are implemented to study the interplay between network structure and stubbornness.
\newpage
\subsection{Resistance parameter}
\label{stubbpar}

@@

We make a fraction of the actors ($f_s$) completely stubborn ($r_i = 1$). We go from no stubborn actors, $f_s=0$, to all the actors are stubborn, $f_s=1$. \\
\newline
First, we discuss the formation of echo chambers for balanced starting conditions, after which the results for the opinion evolution are shared. Then, the formation of echo chambers and the evolution of the prevalence of the opinions for unbalanced starting conditions are discussed.\\
\newline
\textbf{Balanced starting conditions}\\
\newline
We start with a configuration in which $50 \%$ of the actors is given opinion A and $50 \%$ is given opinion B. Fig. \ref{echo_vs_frac_complStubb} shows the formation of echo chambers versus the fraction of stubborn actors, $f_s$. The PR filtering algorithm is used.

>>>>

This chapter is subdivided into four main sections. We first discuss the impact of network structure on the evolution of the prevalence of the opinions and the formation of echo chambers. Then, we introduce stubbornness. The impact of how the opinions are initially distributed in a network with community structure (SBM) are discussed afterwards. Finally, we study the opinion dynamics model in real-world networks. We discuss the results and compare them to results from network models with similar average degree, clustering or modularity.\\
\newline
Four different network models are studied: (i) Erd\H{o}s-R\'{e}nyi model (ER); (ii) Watts-Strogatz model (WS); (iii) stochastic block model (SBM) and (iv) stochastic Watts-Strogatz block model (SBM-WS). Each model has $N = 10^3$ nodes and average degree $\left<k\right> \sim 10$. The ER model has an edge probability $p = 0.01$. For the WS model, we use two rewiring probabilities: $\beta = 0.01$ ($\left<cc\right> = 0.648 \pm 0.003$) and $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). The SBM has 10 communities of size 100 nodes each. We study one configuration with high modularity ($Q = 0.908 \pm 0.005$, with an edge probability inside the communities $p_{\text{cl}} = 0.1$ and an edge probability between communities $p_{\text{add}} = 0.001$) and one configuration with low modularity ($Q = 0.220 \pm 0.007$, with $p_{\text{cl}} = 0.03$ and $p_{\text{add}} = 0.008$). The SBM-WS has 10 communities of size 100 nodes each. Each community has WS parameters: mean degree $\left<k\right> =10$ and rewire probability $\beta = 0.01$. Edges are added between communities with a probability $p_{\text{add}} = 0.001$ ($\left<cc\right> = 0.553 \pm 0.004$ and $Q = 0.909 \pm 0.004$).\\
\newline
The opinion dynamics model discussed in Section \ref{modelThesis} is used. We set the activation probability equal to $p_{\text{act}} = 0.1$ and the actual time-line of each actor has a maximum length equal to $S=20$. If another model is used or if parameters (either those of the network models or the opinion dynamics model) are changed, this will be explicitly stated in the relevant sections. Opinions and stubborns (if present) are randomly distributed over the network, unless specified otherwise.

@@

Real-world social networks often have high clustering and modular structure. We want to study the impact of these topologies on the dynamics of opinions. In this section, we will discuss results related to different network topologies with focus on the effect of clustering and community structure. Two different opinion distributions are investigated, balanced and unbalanced starting conditions. For both starting conditions we investigate the evolution of the prevalence of the opinions and the formation of echo chambers. Two filtering algorithms will be compared, the REC and the PR filtering algorithms, to investigate the interplay between filtering algorithms and network structure.

@@

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_REC_8x7.png}
      \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{}
  	\includegraphics[width=8cm, height=7cm]{echo_chambers_no_stubb_50-50_PR_8x7.png}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_{\text{B}}^{\text{nn}}\right>$, for balanced starting conditions.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$, for balanced starting conditions. \textbf{(a)} REC filtering algorithm; \textbf{(b)} PR filtering algorithm. Four network models are used: (i) ER; (ii) SBM; (iii) SBM-WS and (iv) WS. Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$)) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. For the WS model, we use two different values for the rewire probability: $\beta = 0.01$ ($\left<cc\right> = 0.648 \pm 0.003$; WS2) and $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$; WS1). The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line ($y=1$) represents a constant amount of echo chambers over time. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}
\label{echo_50_50}
\end{figure}
\noindent
Two filtering algorithms are compared, the REC and the PR filtering algorithms. The results for both filtering algorithms are qualitatively the same. The difference is that the formation of echo chambers, using the PR filtering algorithm, is about ten times as large compared to the formation of echo chambers using the REC filtering algorithm. \\
\newline
The ER model shows no, or a limited, sign of the formation of echo chambers, for both the REC and PR filtering algorithms. Networks with high clustering (WS and SBM-WS), on the other hand, show a sign of polarization and formation of echo chambers (visible in the two spikes, one around zero and another around one). Going to higher values of the clustering coefficient ($\left<cc\right> = 0.557 \pm 0.007$ to $\left<cc\right> = 0.648 \pm 0.003$) in the WS model results in significantly larger echo chambers. This suggests that the formation of echo chambers is sensitive to changes in the clustering coefficient. The SBM with high modularity, community structure, shows an appearance of polarization (mainly when using the PR filtering algorithm), however, the formation of echo chambers is significantly smaller compared to networks with high clustering. The stochastic block model with low modularity has no, or limited, formation of echo chambers, similar as for the ER model. This is expected, since a SBM with low modularity, and hence, almost non-existing community structure, resembles an ER network. The SBM-WS has both community structure and high average clustering. The formation of echo chambers when using this model is approximately of the same size as the formation of echo chambers in a WS model with similar average clustering. This suggests that it is the high clustering that drives the dynamics to form echo chambers, while the community structure plays a less important role. \\
\newline
For balanced starting conditions, none of the network models, nor the filtering algorithms could break the status quo (Fig. \ref{ev_op_50_50} and related results in \cite{Perra2019}). Despite this preserving of the balanced opinion distribution, the interplay of filtering algorithms and network structure can shift the initial randomly distributed opinions to a clustered opinion distribution and thus lead to polarization and formation of echo chambers.

@@

\begin{figure}[H]
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_ER_001_50-50_PR_REC_20-80_8x7.png}
      \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
  	\includegraphics[width=8cm, height=7cm]{fraction_of_opinions_WS_10-006_PR_REC_20-80_8x7.png}
    \label{ws_20-80_op}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}    
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM_01-0001_003-0008_PR_REC_20-80_8x7.png}
  \end{subfigure}
  \begin{subfigure}[t]{0.49\textwidth}
    \subcaption{}
    \includegraphics[width=8cm, height=7cm]{fraction_of_opinions_SBM-WS_10-001-0001_PR_REC_20-80_8x7.png}
    \label{sbm-ws_20-80_op}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions.]{Evolution of the prevalence of opinion A ($P_\text{A}(t)$) for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. For each model the REC and the PR filtering algorithms are compared. The dotted black line represents the initial prevalence of opinion A. Each curve is the average of $10^2$ independent simulations and data points are shown for every 50 time steps to improve visualization. Bars represent the standard deviation $\sigma$.}
\label{ev_op_20_80}
\end{figure}
\noindent
This result differs from results in \cite{Perra2019}, where the minority opinion was able to survive over time for all the network models, even for the PR filtering algorithm, and where a minority opinion prevalence at $t=500$ between $P_\text{A}(500) = 0.06$ and $P_\text{A}(500) = 0.09$ was reported \cite{Perra2019}. The difference in the results is most likely due to an interplay between the network size $N$, the average degree $\left<k\right>$ and the maximum length of the actual time-line $S$. The results that where obtained in \cite{Perra2019} are for networks with size $N = 10^5$ nodes and average degree $\left<k\right> \sim 6$. In this thesis results are obtained for $N=10^3$ and average degree $\left<k\right> \sim 10$. In both cases the maximum length of the actual time-line is $S=20$. Fig. \ref{ev_op_WS_k=6} shows the results for the evolution of the prevalence of opinion A for an initial $20/80$ opinion distribution and using the PR filtering algorithm; Fig. \ref{ws1000} displays results for a WS model with $N=10^3$ and $\left<k\right> = 6$; Fig. \ref{ws10000} shows results for a WS model with $N=10^4$ and $\left<k\right> = 6$. We see that the minority opinion A reaches higher final values of $P_0(500)$ compared to the case depicted in Fig. \ref{ws_20-80_op}, with a value of $P_\text{A}(500) = 0.027 \pm 0.023$ in Fig. \ref{ws1000} and a value of $P_\text{A}(500) = 0.030 \pm 0.018$ in Fig. \ref{ws10000}.

@@

  \caption[Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t=500$, $\left<P_\text{B}^{\text{nn}}\right>$, for unbalanced starting conditions.]{Distribution of the fraction of friends (nearest neighbors, nn) of actor $i$ with the same opinion B at $t = 500$ for unbalanced starting conditions. \textbf{(a)} ER; \textbf{(b)} WS; \textbf{(c)} SBM; \textbf{(d)} SBM-WS. The WS model has $\beta = 0.06$ ($\left<cc\right> = 0.557 \pm 0.007$). Two configurations are implemented for the SBM: a high modularity ($Q = 0.908 \pm 0.005$) and a low modularity ($Q = 0.220 \pm 0.007$) configuration. For each model the REC and the PR filtering algorithms are compared. The distribution is normalized by dividing for the same quantity computed at $t=0$. The dotted black line ($y=1$) represents a constant amount of echo chambers over time. Each curve is the average of $10^2$ independent simulations. Error bars are omitted for improved visualization.}

@@

Lets first discuss results for the ER model and the SBM (Fig. \ref{er_echo_20-80} and Fig. \ref{sbm}). When using the REC filtering algorithm, the formation of echo chambers around the majority opinion B is hampered. This is probably a consequence of the absence of clustering and the presence of short average path lengths in these models. There is even a little spike around $\left<P_\text{B}^{\text{nn}}\right> = 0.1$, which indicates a small increase in the fraction of actors with the majority of their neighbors carrying the minority opinion A. From Fig. \ref{ev_op_20_80} it was concluded that the REC filtering algorithm does not break the initial prevalence of the opinions. The opinion distribution is however slightly shifted from the case where the minority opinion A is randomly distributed over the network to the case where the minority opinion A starts to group together in opinion clusters. \\

@@

The PR filtering algorithm shows a different tendency. There is an increase in the fraction of actors with all their neighbors carrying the majority opinion B. This is a consequence from the fact that the PR filtering algorithm reduces the visibility of the minority opinion A in the neighborhood of actors with the majority opinion B. This reduced visibility of the minority opinion A in these neighborhoods drives them eventually to adopt opinion B, with an increase in the number of echo chambers around the majority opinion B as a consequence. Remember from Fig. \ref{ev_op_20_80}, that the PR filtering algorithm reduces the initial prevalence of the minority opinion. Hence, the increase in the formation of echo chambers around the majority opinion B is not only a consequence of a reorganization of the system from a random opinion distribution to a clustered opinion distribution, but also from a shift in the initial $20/80$ opinion distribution to a final (close to) $0/100$ opinion distribution. This can also be seen in the fact that, across the $x$-axis, the average number of neighborhoods with some fraction of the actors carrying the minority opinion A becomes equal to zero. \\

@@

Networks with high clustering (WS and SBM-WS) behave completely different than those with absence of clustering. For both the REC and PR filtering algorithms the high clustering enhances the formation of echo chambers around the minority opinion A (visible in the spike around $\left<P_\text{B}^{\text{nn}}\right> = 0$ in Fig. \ref{ws} and Fig. \ref{sbm-ws}). The absolute value of neighborhoods with all the actors carrying the minority opinion A at $t=0$ is equal to zero in the case of a WS network in Fig. \ref{ws} and in the case of a SBM-WS network in Fig. \ref{sbm-ws}. This leads to a normalized average value of the echo chambers, which is obtained by dividing the value at $t=500$ by the corresponding value at $t=0$, equal to infinity. In Fig. \ref{ws} and Fig. \ref{sbm-ws}, the spikes around zero are not reaching infinity, but are cut off around values equal to 100 or 180. These large spikes, indicating a large increase in echo chambers around the minority opinion, do not necessarily mean that we have a large absolute number of echo chambers around the minority opinion A at $t=500$. They merely indicate that, compared to the beginning at $t=0$, the increase is large (if there are zero echo chambers around the minority opinion A at $t=0$, the increase is infinite even if there is only one echo chamber around the minority opinion A at $t=500$). If we would look at the absolute values of the number of echo chambers at $t=500$, we would still find that the vast majority of the like-minded neighborhoods carry the majority opinion B and that only a minor fraction carries the minority opinion A.\\

@@

\begin{table}[H]
\centering
\setlength{\tabcolsep}{10pt} % Default value: 6pt
\renewcommand{\arraystretch}{1.5} % Default value: 1
\begin{tabular}{c | c}
$\bm{Q}$ & \textbf{Network Model} \\
\hline
$0$ & ER model; $p=0.01$ \\
$0.220 \pm 0.007$ & SBM; $p_{\text{cl}} = 0.03$, $p_{\text{add}} = 0.008$ \\
$0.476 \pm 0.009$ & SBM; $p_{\text{cl}} = 0.05$, $p_{\text{add}} = 0.005$ \\
$0.624 \pm 0.008$ & SBM; $p_{\text{cl}} = 0.07$, $p_{\text{add}} = 0.004$ \\
$0.908 \pm 0.005$ & SBM; $p_{\text{cl}} = 0.1$, $p_{\text{add}} = 0.001$ \\
\end{tabular}
\caption{Network models used in Fig. \ref{echo_vs_mod}.}
\label{tab1}
\end{table}
\noindent
For most values of the modularity, the REC filtering algorithm produces almost no echo chambers. If the modularity becomes high enough we see a small increase in the formation of echo chambers, however even for $Q = 0.908 \pm 0.005$, the size of the echo chambers remains smaller than two. This means that after 500 time steps there are less than twice as much echo chambers than in the beginning at $t=0$. For the PR filtering algorithm, the increase in the size of the echo chambers with increasing modularity is more significant. From a modularity equal to $Q = 0.476 \pm 0.009$, we see a sharp increase in the formation of echo chambers. For $Q < 0.476$, the formation of echo chambers remains small, even when using the PR filtering algorithm. These results suggest that community structure enhances the formation of echo chambers, mostly when using the PR filtering algorithm. High modularity is however required to see the effect of community structure on the formation of opinion clusters.
\newpage
\begin{figure}[H]
	\includegraphics[width=10cm, height=8cm]{echo_chamber_vs_modularity_edge_prob_same_deg_PR_REC_10x8}
	\captionsetup{format=plain}
	\caption[Formation of echo chambers versus community structure (modularity $Q$).]{Formation of echo chambers versus community structure, measured by the modularity $Q$, for both the PR and REC filtering algorithms. Data points with $Q = 0$ are the result for an ER network; the other eight data points are results for SBM's, each with different modularity, which is changed by changing the edge probabilities. The $y$-axis shows the average of the normalized fraction of actors with all neighbors having opinion A and the normalized fraction of actors with all neighbors having opinion B  (normalized means that the value for the fraction of actors with all neighbors having opinion A or all neighbors having opinion B at $t=500$ is divided by the corresponding value at $t=0$). The dotted black line ($y=1$) represents a constant amount of echo chambers over time. Each data point is the average of $10^2$ independent simulations. Bars represent the standard error $SE$. Data points are shown together with the standard error $SE = \sigma /\sqrt{n}$ ($n$ is the number of independent simulations), because we are most interested in the correctness of the average.}
\label{echo_vs_mod}
\end{figure}
\noindent
\section{Stubborn actors}

Stubbornness is an important feature of real-life opinion dynamics. Understanding the impact of individual resistance and the interplay with filtering algorithms and network topologies is crucial. In this section, we will study the effect of stubborn actors on the opinion dynamics. We focus on two different stubbornness models, discussed in Section \ref{stubbBin}. Results of the first model, including a resistance parameter, are given in Section \ref{stubbpar}. Results of the second model, the majority threshold model, are presented in Section \ref{majThres}.\\
\newline
We will compare the REF, REC and PR filtering algorithms to analyze the interplay between stubbornness and filtering algorithm. The ER, WS, SBM and SBM-WS are implemented to study the interplay between network structure and stubbornness.
\newpage
\subsection{Resistance parameter}
\label{stubbpar}

@@

We make a fraction of the actors ($f_s$) completely stubborn ($r_i = 1$). We go from no stubborn actors, $f_s=0$, to all actors are stubborn, $f_s=1$. \\
\newline
First, we discuss the formation of echo chambers for balanced starting conditions, after which the results for the opinion evolution are shared. Then, the formation of echo chambers and the evolution of the prevalence of the opinions for unbalanced starting conditions are discussed.\\
\newline
\textbf{Balanced starting conditions}\\
\newline
We start with a configuration in which $50 \%$ of the actors is given opinion A and $50 \%$ is given opinion B. Fig. \ref{echo_vs_frac_complStubb} shows the formation of echo chambers versus the fraction of stubborn actors, $f_s$. The PR filtering algorithm is used.
